Measurement Clusterization in Bayesian D-optimal Designs in Infinite Dimensions

### 2024-07-22 latex/intro.tex
Compilesgit statusgit status
Fixed intro and prelims

\section{Introduction}\label{section:intro}
The practice of experimental design entails specifying every detail of
running an experiment \cite{chaloner}: at its simplest form,
experimental design involves choosing a set of measurements to be
taken. For example, in a medical experiment, experimental design may
involve choosing participants to whom some drug will be administered,
and measuring their blood pressure. Experimental design also arises in
the process of searching for oil. Since measuremnts are expensive and
require digging holes in the ground, only a limited number of
measuremnts can be taken, hence carefully choosing measuremt locations
is crucial.

Specifiyng the right set of measurements holds particular significance
when solving an \emph{inverse problem} --- i.e. making inference of
the physical world utilizing a physical model a phenomenon of interest
\cite{tarantola,somersalo}. In an inverse problem, we seek to infer
physical quantities, based on observations and a model of the physical
world. Models are typically phrased in the language of ordinary or
partial differential equations.

Inverse problems are ubiquitous in science and engineering, and we
will now give several illustrative examples. In electrical impedance
tomography (EIT) we seek to infer the impedance of the inside of a
human body from observations of electrical current impedance measured
at specific electrode locations on the skin, leveraging known
equations of electric current flow
\cite{horesh2010impedance}. Magnetic Resonance Imaging (MRI) also
entails solving an inverse problem, where radio-frequency pulses are
sent through the human body in the presence of a strong magnetic field
and the response of one's body contents to such perutrbations is
measured, allowing a radiologist to view the internals of a body in a
noninvasive manner \cite{horesh2008mri}. Yet another example of an
inverse problem arises in oil and gas exploration, where acoustic
waves are sent through "boreholes" --- deep cylindrical wells drilled
into the ground --- and data of travel and return times of said waves
are recorded. Travel and return time is influenced by the properties
of the subsurface materials, such as density, elasticity, and the
presence of fluids or voids, combining travel data with a geophisical
models of contents of the earth's crust to facilitate reconstructing
and inferring the structure of the subsurface in a process called
\emph{borehole tomography} \cite{horesh2008borehole}. Inverse problems
also arise in many other areas of seismology and geology
\cite{rabinowitz1990, steinberg1995} and medical imaging
\cite{tarantola}. While in many inverse problems the goal is to infer
some parameter, in all of the above examples and in the formulation we
consider here, the goal is to infer some \emph{function} over $
\Omega$, where (usually) \(\Omega \subseteq \mathbb{R}^d, d=1,2,3\) is
the spatial domain of interest.

%% In EIT, electrode locations determine the quality of the
%% reconstructed impedance field, while in MRI utilizng certain
%% radio-frequency pulses may give more accurate information than
%% others. In borehole tomogrpahy, only several boreholes could be dug
%% and the choice of their locations is also crucial for oil and gas
%% discovery. Thus, the selection of optimal measurements, referred to as
%% the problem of \emph{optimal design}, becomes crucial.
%% in a specific location (in a typical
%% medical application to measure inverventions on, . These experimental
%% units may be individual patients, but also Although common sense is
%% useful in this process, it is often the case that naive intuition
%% results in experiments that do not generate the maximal amount of
%% information possible. To this end

For concreteness, let us consider a toy model of an inverse problem:
the \emph{heat equation} in one dimension (1D heat equation
henceforth, see Section \ref{subsec:1d heat} for more details). The
heat equation is a partial differential equation compose of three
equations: eqs.~\eqref{eq:heat1}, \eqref{eq:heat2}, and
\eqref{eq:heat3}. Readers less familiar with partial differential
equations should think of the heat equation as follows: before us
there is a perfectly insulated metal rod. At each point on the rod the
temperature is different and unknown, and we call this temperature
distribution the \emph{initial condition}. We wait for a short time
$T$ and let the heat dissipate a bit inside the rod. The final heat
distribution in the rod is dictated by the following set of equations
which defines the 1D heat equation:
\begin{subequations}%\label{eq:heat equation}
  \begin{alignat}{2}
    u_t &= \alpha \Delta u &&\qquad \text{in } [0,1] \times [0,\infty), \label{eq:heat1}\\
    u &= 0 &&\qquad \text{on } \{0, 1\} \times [0,\infty), \label{eq:heat2}\\
    u &= u_0 &&\qquad \text{on }[0,1] \times \{0\}. \label{eq:heat3}\\
  \end{alignat}
\end{subequations}

As time passes, heat disspates across the rod as hotter regions,
e.g.~local maximas of the heat distribution, become cooler. This
behavior is captured by eq.~\eqref{eq:heat1}: the Laplacian at local
maxima is negative, so $u_t = \Delta u$ implies $u$ decreases at its
maximas, and the reverse happens at local minima. Eq.~\eqref{eq:heat2}
describes the \emph{boundary condition}, i.e.~the interaction of the
heat inside the rod with the boundary of the domain $\Omega =
[0,1]$. Here we consider a \emph{homogeneous Dirichlet boundary
condition} which is best understood as an absorbing boundary: heat
that interacts with the boundary is immediately
absorbed. Eq.~\eqref{eq:heat3} describes the \emph{initial condition}
of the heat, i.e.~the heat distribution at time $t=0$. Thus, we have
described the \emph{forward} problem: how heat evolves in time,
starting at $t=0$ up until $t=T$.

In the inverse problem of the 1D heat equation, our goal is to infer
the initial condition $u_0 = u(\cdot, 0)$ from observations of the
final condition $u(\cdot, T)$. We are not able to measure $u(\cdot,
T)$ at every point in the domain $\Omega$. Rather, we take some number
of temperature measurements on the rod once our waiting time $T$ has
passed and try to infer $u(\cdot, 0)$ from these measurements.

The inverse problem of the 1D heat equation is difficult ("ill-posed")
since heat spreads in a diffusive manner and any roughness in the
initial condition is quickly smoothed, see Supplementary movies S1 and
S2. For ill-posed problems regularization is in order, and we
implement such regularization via a fully Bayesian formulation of the
inverse problem (see Section \ref{section:prelims} for details).

In all of the examples of inverse problems we have seen, as well as in
many other applications, measurements should be chosen to most benefit
the user. Hence, the user may consider some utility, called a
\emph{design criterion} and take measurements that maximize this
utility. Two of the most widely recognized and extensively studied
design criteria are A- and D-optimality. We focus on the Bayesian
D-optimality criterion, which carries a simple and intuitive meaning:
measurements chosen according to the D-optimality criterion maximize
the expected Kullback-Liebler divergence from posterior to prior
\cite{Chaloner1995, AlexanderianGloorGhattas14}, see Section
\ref{subsec:D optimal design} for a precise definition. We refer to a
set of measuremnts that maximizes the Bayesian D-optimality criterion
a \emph{D-optimal design}.

Surprisingly, A- and D-optimal designs for inverse problems have been
observed to yield remarkably similar measurements in certain cases
\cite{fedorov1996, nyberg2012, fedorov1997, Ucinski05,
  neitzel2019sparse}. This phenomenon is illustrated for our toy
inverse problem in Fig.~\ref{fig:clusterization illustration}, where
D-optimal measurement locations are shown for different numbers of
measurements. Notably, for six measurements, a D-optimal design yields
two sets of measurements that are almost indistinguishable from one
another. We refer to this intriguing phenomenon as \emph{measurement
clusterization} \cite{Ucinski05}. We consider a design to be
\emph{clustered} when two or more of its constituent measurements are
identical.

@@

\begin{enumerate}
\item \label{q:generic} \textbf{Is measurement clusterization a
  generic phenomenon?}
  %\subsection{An answer for Question \ref{q:generic}: Genericity of measurement clusterization}
  %% Computer implementation of Lemma \ref{thm:char} for the inverse
  %% problem outlined in Section \ref{section:how} also generates \(\obs\)
  %% as a solution to the D-optimal design problem.  Furthermore,
  Randomized numerical simulations of our model give rise to D-optimal
  designs that exhibit clusterization more than 95\% of the time (see
  Section \ref{subsec:lemma sims} for details). Given the genericity
  of our model, we expect measurement clusterization to be a generic
  and ubiquitous phenomenon.

@@

The implication of our answer to Question \ref{q:generic} is that
encountering clusterization should be expected in many different
inverse problems across many different scientific fields and
disciplines. Thus, researchers that encounter clusterization should
not be surprised or wary. Then, in our answer to Question \ref{q:why},
we explain what is the cause of clusterization. As it appears, the
cause is universal: a D-optimal design reduces uncertainty for a
select set of prior covariance eigenvectors --- those with the most
prior uncertainty, i.e.~those the practitioner cares about the most!
Consequently, practitioners should not try to avoid measurement
clusterization. Rather, practitioners should should take repeated
measurements (e.g.~in MRI and borehole tomography), increase apparatus
sensitivity (e.g.~in EIT), or take consecutive measurements (e.g.~in
the 1D heat equation). Overall, we show that measurement
clusterization is a natural and (almost) inevitable part of Bayesian
D-optimal designs.

\subsubsection{Further Contributions}
In Theorem \ref{thm:char}, we also show that D-optimal designs are
best understood in the sapce of \emph{observations}. Again, a precise
statement could be found in Section \ref{section:vanishing}.  This
result is in accordance with previous work by \cite{koval2020}, who
showed that A-optimal designs are best constructed in the space of
observations.

@@

Finally, we must stress that we do not suggest that clustered designs
should be abandoned, nor do we support the use of clustered designs in
any circumstances. We merely investigate an intriguing phenomenon of
clusterization and seek a deeper understanding for it. Researchers in
other disciplinesis not viewed as a problem. For example, \cite{lozan}
suggest in the introduction section that repeated measurements are
necessary. Our results align with that common wisdom and settle the
seemed contradiction.

>>>>

\section{Introduction}\label{section:intro}
The practice of experimental design entails specifying every detail of
running an experiment \cite{chaloner1995}: at its simplest form,
experimental design involves choosing a set of measurements to be
taken. For example, consider the process of searching for oil:
measuremnts are expensive and require digging holes in the ground, so
only a limited number of measuremnts can be taken, hence carefully
choosing measuremt locations is crucial.

Specifiyng the right set of measurements holds particular significance
when solving an \emph{inverse problem} --- i.e.~making inference of
the physical world utilizing a physical model a phenomenon of interest
\cite{tarantola2005,kaipio2005}. In an inverse problem, we seek to
infer physical quantities, based on observations and a model of the
physical world. Models are typically phrased in the language of
ordinary or partial differential equations.

Inverse problems are ubiquitous in science and engineering. For
example, in electrical impedance tomography (EIT) we seek to infer the
structure of the inside of a human body. Inference is conducted based
on observations of electrical current impedance measured at specific
electrode locations on the skin, leveraging known equations of
electric current flow \cite{horesh2010impedance}. Magnetic Resonance
Imaging (MRI) also entails solving an inverse problem. There,
radio-frequency pulses are sent through the human body in the presence
of a strong magnetic field. The response of one's body contents to
these radio-frequency pulses is measured, allowing a radiologist to
view the internals of a body in a noninvasive manner
\cite{horesh2008mri}. An inverse problem also arises in oil and gas
exploration. Acoustic waves are sent through "boreholes" --- deep
cylindrical wells drilled into the ground. Data of travel and return
times of the acoustic waves are recorded. Wave travel and return time
is influenced by the properties of the subsurface materials, such as
density, elasticity, and the presence of fluids or voids. Combining
travel data with a geophisical models of contents of the earth's crust
facilitates reconstructing and inferring the structure of the
subsurface in a process called \emph{borehole tomography}
\cite{horesh2008borehole}. Inverse problems also arise in many other
areas of seismology and geology \cite{rabinowitz1990, steinberg1995}
and medical imaging \cite{tarantola2005}. While in many inverse
problems the goal is to infer some parameter, in all of the above
examples and in the formulation we consider in this study, the goal is
to infer some \emph{function} over $ \Omega$, where (usually) \(\Omega
\subseteq \mathbb{R}^d, d=1,2,3\) is a spatial domain of interest.

\subsection{D-optimal Designs}\label{subsec:D}
In all of the examples of inverse problems we have seen, as well as in
many other applications, only a limited number of measurements are
allowed. For example, in borehole tomography, a measurement involves
drilling a deep well in the ground --- a very costly endeavor. Even in
a seemingly simpler application such as MRI, a patient cannot occupy
the MRI machine for too long, hence only a limited number of
measurements can be taken. Therefore, measurements should be chosen to
most benefit the user. Typically, a user will consider some utility,
called a \emph{design criterion} and take measurements that maximize
this utility. Two of the most widely recognized and extensively
studied design criteria are A- and D-optimality.

In this study, we focus on the Bayesian D-optimality
criterion. Bayesian D-optimality carries a simple and intuitive
meaning: measurements chosen according to the D-optimality criterion
maximize the expected Kullback-Liebler divergence from posterior to
prior \cite{chaloner1995, AlexanderianGloorGhattas14,
  CoverThomas91}. Utilizing KL divergence for quantifying distances
between distributions is standard practice and we will not motivate it
further here. For a discrete parameter $\param$ and data $\data$, the
KL divergence is defined as
\begin{equation}\label{eq:basic KL}
  D_{KL}\left (\Pr(\param|\data)||\Pr(\param)\right ) =  \sum_{\param} \log
  \frac{\Pr(\param|\data)}{\Pr(\param)} \Pr(\param|\data). 
\end{equation}
Of course, data $\data$ is not known before the experiment is
conducted, so we average over it to define the D-optimality criterion
as:
\begin{equation*}
  \mathbb{E}_{\data}\left [ D_{KL}\left (\Pr(\param|\data)||\Pr(\param)\right ) \right ]
\end{equation*}
We refer to a set of measuremnts that maximizes the D-optimality
criterion as a \emph{D-optimal design}.

For a linear model with Gaussian prior in finite dimensions, a
D-optimal design minimizes the determinant of the posterior covariance
\cite{chaloner1995}. In Section \ref{subsec:D optimal design} we give
a more general definition of the D-optimality criterion that applies
to arbitrary measures.

\subsection{A toy model: the 1D heat equation}
For concreteness, let us now consider a toy model of a physical process
for which we wish to solve an inverse problem: a partial differential
equation known as the \emph{heat equation} in one dimension (1D heat
equation henceforth, see details below). Readers less familiar with
partial differential equations can think of the following setup:
before us there is a perfectly insulated metal rod. At each point on
the rod the temperature is different and unknown, and we call this
temperature distribution the \emph{initial condition}. We wait for a
short time $T$ and let the heat dissipate a bit inside the rod. The
heat equation determines the heat distribution inside the rod at time
$t=T$.

The full time evolution of heat in the rod $\Omega=[0,1]$ is formally
described by the following three equations:
\begin{subequations}%\label{eq:heat equation}
  \begin{alignat}{2}
    u_t &= \Delta u &&\qquad \text{in } [0,1] \times [0,\infty), \label{eq:heat1}\\
    u &= 0 &&\qquad \text{on } \{0, 1\} \times [0,\infty), \label{eq:heat2}\\
    u &= u_0 &&\qquad \text{on }[0,1] \times \{0\}. \label{eq:heat3}
  \end{alignat}
\end{subequations}
As time passes, heat disspates across the rod as hotter regions,
e.g.~local maximas of the heat distribution, become cooler. This
behavior is captured by eq.~\eqref{eq:heat1}: the Laplacian at local
maxima is negative, so $u_t = \Delta u$ implies $u$ decreases at its
maximas, and the reverse happens at local minima. Eq.~\eqref{eq:heat2}
describes the \emph{boundary condition}, which dictates how heat
interacts with the outside world at the two edges of the rod
$\{0,1\}$. Specifically, eq.~\eqref{eq:heat2} implements an absorbing
boundary, i.e.~heat that interacts with the boundary is immediately
absorbed and disappears. This type of boundary condition is known as a
\emph{homogeneous Dirichlet boundary condition}. Eq.~\eqref{eq:heat3}
describes the \emph{initial conditions}, i.e.~the heat distribution
inside the rod at time $t=0$.

In the \emph{inverse problem of the 1D heat equation}, our goal is to
infer the initial condition $u_0 = u(\cdot, 0)$ from observations of
the final condition $u(\cdot, T)$. However, we are not able to measure
$u(\cdot, T)$ at every point in the domain $\Omega$. Rather, we take
some number of temperature measurements on the rod once our waiting
time $T$ has passed and try to infer $u(\cdot, 0)$ from these
measurements.

The inverse problem of the 1D heat equation is difficult ("ill-posed")
since heat spreads in a diffusive manner and any roughness in the
initial condition is quickly smoothed, see Supplementary movies S1 and
S2. For ill-posed problems regularization is in order, and we
implement such regularization via a fully Bayesian formulation of the
inverse problem, see Section \ref{section:prelims} for details.

\subsection{Measurement Clusterization}
Surprisingly, A- and D-optimal designs for inverse problems have been
observed to yield remarkably similar measurements in certain cases
\cite{fedorov1996, nyberg2012, fedorov1997, Ucinski05,
  neitzel2019sparse}. This phenomenon is illustrated for our toy
inverse problem in Fig.~\ref{fig:clusterization illustration}, where
D-optimal measurement locations are shown for different numbers of
measurements. Notably, for six measurements, a D-optimal design yields
two sets of measurements that are almost indistinguishable from one
another. We refer to this intriguing phenomenon as \emph{measurement
clusterization} \cite{Ucinski05}. We consider a design to be
\emph{clustered} when two or more of its constituent measurements are
identical.

@@

\begin{enumerate}
\item \label{q:generic} \textbf{Is measurement clusterization a
  generic phenomenon?}
  %\subsection{An answer for Question \ref{q:generic}: Genericity of measurement clusterization}
  %% Computer implementation of Lemma \ref{thm:char} for the inverse
  %% problem outlined in Section \ref{section:how} also generates \(\obs\)
  %% as a solution to the D-optimal design problem.  Furthermore,
  We give two complementing answers to this question. First, from a
  theoretical perspective, we show that clusterization barely depends
  on the specifics of the prior, so long as it is Gaussian. We show
  that the structure of the decay of the eigenvalues of the
  pushforward prior covariance determine clusterization. The decay of
  these eigenvalues depends on how ill-posed the problem at hand is
  and does not depend much on the prior. see Theorem \ref{thm:char}
  and the Discussion for details. From a numerical experiments
  perspective, simulations of our model give rise to D-optimal designs
  that exhibit clusterization more than 95\% of the time (see Section
  \ref{subsec:lemma sims} for details). Given the genericity of our
  model, we expect measurement clusterization to be a generic and
  ubiquitous phenomenon.

@@

Our results have several implications. Our answer to Question
\ref{q:generic} implies that encountering clusterization should be
expected in many different problems across many different scientific
fields. Thus, researchers that encounter clusterization should not be
surprised or wary. In Our answer to Question \ref{q:why}, we explain
what is the cause of clusterization. As it appears, the cause is
generic: a D-optimal design reduces uncertainty for a select set of
prior covariance eigenvectors --- those with the most prior
uncertainty, i.e.~those the practitioner cares about the most!
Consequently, practitioners should not try to avoid measurement
clusterization. Rather, practitioners should take repeated
measurements (e.g.~in MRI and borehole tomography), increase apparatus
sensitivity (e.g.~in EIT), or take consecutive measurements (e.g.~in
the 1D heat equation). Overall, we show that measurement
clusterization is a natural and (almost) inevitable part of Bayesian
D-optimal designs.

\subsubsection{Further Contributions}
In Theorem \ref{thm:char}, we also show that D-optimal designs are
best understood in the sapce of \emph{observations}, see Section
\ref{section:vanishing} for a precise statement. This is in accordance
with previous work by \cite{koval2020}, who showed that A-optimal
designs are best constructed in the space of observations.

@@

%% Finally, we must stress that we do not advocate any particular
%% solution to clusterization. We do not suggest that clustered designs
%% should be abandoned, nor do we support the use of clustered designs in
%% any circumstances. Our goal is to investigate an intriguing phenomenon
%% of clusterization and seek a deeper understanding for it. Different
%% researchers in different disciplines have differenot viewed as a problem. For
%% example, \cite{lozan} suggest in the introduction section that
%% repeated measurements are necessary. Our results align with that
%% common wisdom and settle the seemed contradiction.

### 2024-07-22 latex/prelims.tex
before diff
Fixed intro and prelims

\section{Preliminaries and Notation}\label{section:prelim}
Here we present the basics of Bayesian inverse problems over hilbert
spaces. Since we are ultimately interested in inferring a function over some
domain, we will keep in mind that these Hilbert spaces should really
be thought of as function spaces. The theoretical foundations for
inverse problems over function spaces can be found in \cite{Stuart10}
and will be only briefly reviewed here.

\subsection{Forward Problems}\label{subsec:abstract OED}
In this section we give definitions and notations for forward probles,
which are an essential part of the inverse problems we discuss
later. Consider a "parameter space" \(\hilp\) and an "observation
space" \(\hilo\) --- both separable Hilbert spaces (the subscripts p
and o are for "parameter" and "observation", respectively). The
parameter space includes the quantity we seek to infer; in the inverse
problem of the heat equation, the parameter space $\hilp$ is where the
initial condition lives. The observation space $\hilo$, on the other
hand, is the space from which we take measurements; in the example of
the 1D heat equation, $u(\cdot, T) \in \hilo$.

The connection between parameter and observation spaces is made by the
\emph{forward operator} \(\fwd: \hilp \to \hilo\). We assume the
forward operator \(\fwd\) is linear. In the inverse problem of the 1D
heat, the forward operator is determined by the time evolution of the
1D heat equation \eqref{eq:heat1}, \eqref{eq:heat2} and
\eqref{eq:heat3}.

Measurements are taken via a \emph{measurement operator} \(\obs\). It
is common for the measurement and forward operators to be merged
\(\tmp := \obs \fwd\) \cite{AlexanderianGloorGhattas14}, but the
analysis carried out in the following sections requires that \(\fwd\)
and \(\obs\) are explicitly separated (as in
\cite{attia2022stochastic, cvetkovic2023choosing}). Note that \(\obs
\in ( \hilo^* )^m\), where \(m\) is the number of measurements
taken\footnote{The alert reader will likely ask how do we reconcile
point measurements $\delta_x$ as suggested by the formulation of the
1D heat equation. We don't. This entire study could be formulated for
Banach spaces where point measurements are allowed. We avoid this
technical difficulty because: (1) it does not add much to the study,
(2) it is beyond the technical abilities of the author}. Entries
\(\meas_j, j=1,\dots,m\) of the measurement operator \(\obs\) are
called \emph{measurements}:

@@

Data is acquired via noisy observations, and we consider two types of
error terms: Spatially correlated model error \(\eps' \sim
\normal(0,\modcov)\) with \(\modcov\) a covariance operator. Observation
error is denoted \(\eps \sim \normal(0, \sigma^2 I_m)\), with \(I_m \in
\mathbb{R}^{m \times m}\) the identity. Both error terms and the prior
are assumed independent of each other. Thus, data is acquired via
\begin{align}\label{eq:inverse problem}
  \data := \obs (\fwd \param + \eps') + \eps = \obs \fwd \param + \obs \eps' + \eps.
\end{align}

@@

\begin{align}\label{eq:Sigma}
  \begin{split}
    \Sigma(\obs) :&= \mathbb{E}[ (\obs \eps' + \eps) (\obs \eps' +
      \eps)^t ]
    % 
    % 
    = \obs \modcov \obs^* + \sigma^2I_m , 
  \end{split}
\end{align}
where
\begin{align}\label{eq:modcov explained}
  \begin{split}
    [\obs \modcov \obs^*]_{ij} & = e_i^t \obs \modcov \obs^* e_j 
    %
    %
    %
    = \meas_i (\modcov \meas_j).% \text{ (by \eqref{eq:obs*})}.
  \end{split}
\end{align}
Taking \(\modcov = 0\) is a common practice
\cite{tarantola2005,Kaipio2005,Vogel02} and then \(\Sigma =
\sigma^2I_m\) is a scalar matrix which does not depend on \(\obs\).

\subsection{Bayesian Linear Inverse Problems}\label{subsec:bayesian inverse problems}
In the previous section, we saw how a parameter $u\in \hilp$ is
transported to the observation space via the forward operaror $\fwd u
\in \hilo$, how observations are generated via $\obs \fwd u$ and how
observations and noise give rise to data $\data$. It is time to
formulate the process of inferring the parameter as a Bayesian inverse
problem. The first ingredient we need is a prior.

We take a Gaussian prior \(\param \sim \pr = \normal(\prmean
,\prcov)\) with some appropriate covariance operator \(\prcov\) on
\(\hilp\) \cite{Stuart10}. For the inverse problem of the 1D heat
equation we choose $\prcov = -\Delta^{-1.1}$, see Section
\ref{subsec:1d heat} for details and a short discussion of
"Laplacian-like" covariance operators. Note that \(\fwd \prcov
\fwd^*\) is the prior covariance in \(\hilo\) \cite{Stuart10}, and as
such, \(\fwd \prcov \fwd^*\) is assumed invertible --- an assumption
which will be used below (if \(\fwd\) has a nontrivial kernel we
utilize Occam's Razor and ignore said kernel altogether).

Since this inverse problem is linear and the prior is Gaussian --- the
posterior measure \(\post\) is Gaussian as well. We do not utilize the
posterior mean in this study and posterior covariance operator
$\postcov$ is given by \cite{Stuart10}:
\begin{align}\label{eq:postcov}
  \postcov = (\prcov^{-1} + \fwd^* \obs^* \Sigma^{-1} \obs \fwd
  )^{-1}.
\end{align}

\subsection{Bayesian D-Optimal Designs}\label{subsec:D optimal design} 
A Bayesian D-optimal design maximizes expected Kullback-Leibler (KL)
divergence between posterior \(\post\) and prior measures \(\pr\). It is
useful to first recall the definition of KL divergence \cite{CoverThomas91}:
\begin{equation*}
  D_{KL}(\post||\pr) = \int \log \frac{\der \post}{\der \pr}(\param) \der \post(\param).
\end{equation*}

@@

Intuition for Theorem \ref{thm:d optimality} can be gained by
considering a Bayesian linear model in finite dimensions, with
Gaussian prior and Gaussian noise. Then, a D-optimal design minimizes
the determinant of the posterior covariance matrix, and this turns out
to be a regularized version of the frequentist D-optimality criterion
\cite{Chaloner1995}. Theorem \ref{thm:d optimality} and Definition
\ref{def:d optimality} carry a similar intuition:
\begin{align*}
  \begin{split}
    \tar(\obs) &= \frac12 \log \det ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}) \text{ by \eqref{eq:objective}}\\
    &= \frac12 \log \det \Big( \prcov ( \prcov^{-1} + \fwd ^* \obs^* \Sigma^{-1} \obs \fwd) \Big )\\
    &= \frac12 \log \det \prcov \postcov^{-1} \text{ by \eqref{eq:postcov}}.
    %%  &= \frac12 \log \det \prcov -\frac12 \log \det \postcov.
  \end{split}
\end{align*}
We think of \(\prcov\) as constant, so a D-optimal design minimizes a
quantity analogous to the posterior covariance determinant, similarly
to the finite-dimensional case.

>>>>

\section{Preliminaries and Notation}\label{section:prelim}
Here we present the basics of Bayesian inverse problems over hilbert
spaces. Since we are ultimately interested in inferring a function
over some domain, we will keep in mind that these Hilbert spaces
should really be thought of as function spaces. A deeper treatment of
the foundations for inverse problems over function spaces can be found
in \cite{Stuart10}.

\subsection{Forward Problems}\label{subsec:abstract OED}
In this section we give definitions and notations for forward problems,
which are an essential part of the inverse problems we discuss
later. Consider a "parameter space" \(\hilp\) and an "observation
space" \(\hilo\) --- both separable Hilbert spaces (the subscripts p
and o are for "parameter" and "observation", respectively). The
parameter space includes the quantity we seek to infer; in the inverse
problem of the heat equation, the parameter space $\hilp$ is where the
initial condition lives. The observation space $\hilo$, on the other
hand, is the space from which we take measurements; in the example of
the 1D heat equation, $u(\cdot, T) \in \hilo$.

The connection between parameter and observation spaces is made by the
\emph{forward operator} \(\fwd: \hilp \to \hilo\). We assume the
forward operator \(\fwd\) is linear. In the inverse problem of the 1D
heat, the forward operator is determined by the time evolution of the
1D heat equation \eqref{eq:heat1} and \eqref{eq:heat2}.

Measurements are taken via a \emph{measurement operator} \(\obs\). It
is common for the measurement and forward operators to be merged
\(\tmp := \obs \fwd\) \cite{AlexanderianGloorGhattas14}, but the
analysis carried out in the following sections requires that \(\fwd\)
and \(\obs\) are explicitly separated (as in
\cite{attia2022stochastic, cvetkovic2023choosing}). Note that \(\obs
\in ( \hilo^* )^m\), where \(m\) is the number of measurements
taken\footnote{The alert reader will likely ask how do we reconcile
point measurements $\delta_x$ as suggested by the formulation of the
1D heat equation with workin in Hilbert spaces. We don't. This entire
study could be formulated for Banach spaces where point measurements
are allowed. We avoid this technical difficulty because: (1) it does
not add much to the study, (2) it is beyond the technical abilities of
the author}. Entries \(\meas_j, j=1,\dots,m\) of the measurement
operator \(\obs\) are called \emph{measurements}:

@@

Data is acquired via noisy observations, and we consider two types of
error terms: Spatially correlated model error \(\eps' \sim
\normal(0,\modcov)\) with \(\modcov\) a covariance operator; and
observation error denoted \(\eps \sim \normal(0, \sigma^2 I_m)\), with
\(I_m \in \mathbb{R}^{m \times m}\) the identity. Both error terms and
the prior (see below) are assumed independent of each other. Thus,
data is acquired via
\begin{align}\label{eq:inverse problem}
  \data := \obs (\fwd \param + \eps') + \eps = \obs \fwd \param + \obs \eps' + \eps.
\end{align}

@@

\begin{align}\label{eq:Sigma}
  \begin{split}
    \Sigma(\obs) :&= \mathbb{E}[ (\obs \eps' + \eps) (\obs \eps' +
      \eps)^t ]
    % 
    % 
    = \obs \modcov \obs^* + \sigma^2I_m , 
  \end{split}
\end{align}
where
\begin{align}\label{eq:modcov explained}
  \begin{split}
    [\obs \modcov \obs^*]_{ij} & = e_i^t \obs \modcov \obs^* e_j 
    %
    %
    %
    = \meas_i (\modcov \meas_j).% \text{ (by \eqref{eq:obs*})}.
  \end{split}
\end{align}
Taking \(\modcov = 0\) is a common practice
\cite{tarantola2005,kaipio2005,Vogel02} and then \(\Sigma =
\sigma^2I_m\) is a scalar matrix which does not depend on \(\obs\).

\subsection{Bayesian Linear Inverse Problems}\label{subsec:bayesian inverse problems}
In the previous section, we saw how a parameter $u\in \hilp$ is
transported to the observation space via the forward operaror $\fwd u
\in \hilo$, how observations are generated from a parameter via $\obs
\fwd u$ and how observations and noise give rise to data $\data$. It
is time to formulate the process of inferring the parameter as a
Bayesian inverse problem. The first ingredient we need is a prior.

We take a Gaussian prior \(\param \sim \pr = \normal(\prmean
,\prcov)\) with some appropriate covariance operator \(\prcov\) on
\(\hilp\) \cite{Stuart10}. For the inverse problem of the 1D heat
equation we choose $\prcov = -\Delta^{-1}$ --- see Section
\ref{subsec:1d heat} for details and a short discussion of
"Laplacian-like" covariance operators. Note that \(\fwd \prcov
\fwd^*\) is the prior covariance in \(\hilo\) \cite{Stuart10}, and as
such is assumed invertible --- an assumption which we will use later
(if \(\fwd\) has a nontrivial kernel we utilize Occam's Razor and
ignore said kernel altogether).

Since $\fwd$ is linear and $\pr$ is Gaussian --- the posterior
\(\post\) is Gaussian as well. We do not utilize the posterior mean in
this study, but the posterior covariance operator $\postcov$ is given
by \cite{Stuart10}:
\begin{align}\label{eq:postcov}
  \postcov = (\prcov^{-1} + \fwd^* \obs^* \Sigma^{-1} \obs \fwd
  )^{-1}.
\end{align}

\subsection{Bayesian D-Optimal Designs}\label{subsec:D optimal design} 
A Bayesian D-optimal design maximizes the expected Kullback-Leibler
(KL) divergence between posterior \(\post\) and prior measures
\(\pr\). For arbitrary posterior and prior measures, the KL divergence
is defined, analogously to eq.~\ref{eq:basic KL}, via the
Radon-Nikodym derivative:
\begin{equation*}
  D_{KL}(\post||\pr) = \int \log \frac{\der \post}{\der \pr}(\param) \der \post(\param).
\end{equation*}

@@

Intuition for Theorem \ref{thm:d optimality} can be gained by
recalling from Section \ref{subsec:D} that for a Bayesian linear model
in finite dimensions, with Gaussian prior and Gaussian noise, a
D-optimal design minimizes the determinant of the posterior covariance
matrix. Theorem
\ref{thm:d optimality} and Definition \ref{def:d optimality} carry a
similar intuition:
\begin{align*}
  \begin{split}
    \tar(\obs) &= \frac12 \log \det ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}) \text{ by \eqref{eq:objective}}\\
    &= \frac12 \log \det \Big( \prcov ( \prcov^{-1} + \fwd ^* \obs^* \Sigma^{-1} \obs \fwd) \Big )\\
    &= \frac12 \log \det \prcov \postcov^{-1} \text{ by \eqref{eq:postcov}}.
    %%  &= \frac12 \log \det \prcov -\frac12 \log \det \postcov.
  \end{split}
\end{align*}
We think of \(\prcov\) as constant, in the sense that $\prcov$ does
not depend on data $\data$. Thus, a D-optimal design minimizes a
quantity analogous to the posterior covariance determinant, similarly
to the finite-dimensional case.
